---
title: "Compare IHME to observed counts"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compare IHME to observed counts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(covidutilities)
library(magrittr)
library(ggplot2)
```

IHME has prepared a series of projections dating from March 26 through
April 29. We compared these projections to the observed counts reported
by the Johns Hopkins University Center for Systems Science and Engineering (JHUCSSE)
on their github site (https://github.com/CSSEGISandData/COVID-19).
We obtained the IHME projections from their website (http://www.healthdata.org/covid/data-downloads). 
We decided to compare the IHME projections to the JHUCSSE observed
counts because the IHME documentation reports that IHME used those
counts as their data source for US data. However, for those dates 
on which the IHME files contain observed daily counts, we found that those
counts are different from the daily counts in the JHUCSSE files.
However, the counts add up to approximately the same totals.

The IHME datasets have a row for each day from either January 3, 2020 or February 6, 2020, up to August
4, 2020. They contain a column labeled `deaths_mean` which has the observed death counts up to the
date the projection was made, and the projected death counts thereafter. In order to make the graphs
neater, we remove the observed death counts from this column by setting them to missing, leaving
only the projected death counts. To help with this, the function `ihme_observed_date` adds a column
with the date of the last observed death count. 

```{r projVsObserved2, eval = T, message = F, warning = F}
## Download the observed death and case counts from the JHUCSSE github site,
## and extract the Washington data.
jhu <- jhu_get_data() %>%
    jhu_extractWA() %>%
    dplyr::select(state, date, deaths_day)

## Download the IHME projections and combine them in one file
all_ihme <- NULL
for(proj in c("March 26", "March 27", "March 29", "March 30", "March 31",
              "April 1", "April 5", "April 7", "April 8", "April 10", "April 13",
              "April 17", "April 21", "April 22", "April 27", "April 28", "April 29",
              "May 4", "May 10")){
    all_ihme <- ihme_download(proj) %>%
        ihme_extractUS() %>%
        ihme_extractWA() %>%
        ihme_observed_date() %>%
        dplyr::mutate(deaths_mean = ifelse(date > last_observed_date, deaths_mean, NA),
               deaths_upper = ifelse(date > last_observed_date, deaths_upper, NA),
               deaths_lower = ifelse(date > last_observed_date, deaths_lower, NA)) %>%
        dplyr::select(state, date, deaths_mean, deaths_upper, deaths_lower, projection_date,
                      last_observed_date) %>%
        dplyr::bind_rows(all_ihme)    
}

## Join the IHME and JHU data, and add a date_order column to help sort the
## facets in the graph below.
all_ihme2 <- jhu %>%
    dplyr::left_join(all_ihme, by = c('state','date')) %>%
    dplyr::filter(date >= "2020-03-01") %>%
    dplyr::mutate(date_order = as.Date(projection_date, "%B %d"))

## Prepare a data frame with the strip labels and their position
maxX <- with(all_ihme2, max(max(deaths_upper, deaths_day, na.rm = T)))
proj_dates <- all_ihme2 %>%
    dplyr::distinct(projection_date, .keep_all = T) %>%
    dplyr::mutate(deaths_mean = maxX)

```

The figure below shows the daily observed death counts from the Johns Hopkins University github site in black,
and the projected counts from IHME in red, for each of the IHME projections through May 10.
The black curves are smoothed lines through the observed counts.
The red shaded areas show IHME's prediction intervals.

```{r plotproj2, eval = T, echo = F, message = F, fig.width = 10, fig.height = 12, warning = F}
p3 <- ggplot(all_ihme2, aes(x = date, y = deaths_mean)) +
    geom_line(color = 'red', lwd = 1) +
    geom_ribbon(aes(ymin = deaths_lower, ymax = deaths_upper), fill = "red", alpha = 0.3) +
    geom_smooth(aes(y = deaths_day), lwd = 1, se = F, color = "black") +
    geom_point(aes(y = deaths_day), size = 1) +
    geom_text(data = proj_dates, 
        ##       mapping = aes(label = projection_date), 
              aes(label = projection_date), 
              vjust = "inward", 
              hjust = "inward",
              fontface = "bold", 
              color = "black", 
              size = 3.5) +
    facet_wrap(~reorder(projection_date, date_order), ncol = 3) +
    theme_bw() +
    labs(y = "Deaths per day") +
    scale_x_date(name = "Date",
                 date_breaks = "2 weeks",
                 date_labels = "%b %d") +
    theme(
        ## turn off the strip label and tighten the panel spacing
        strip.text = element_blank(),
        panel.spacing.x = unit(-0.05, "lines"),
        panel.spacing.y = unit(0.3, "lines")
        )
p3

```

